---
title: Shocks
author: "Kiara Hernandez"
date: \today
institute: Harvard University
fontsize: 10pt
output:
 beamer_presentation:
    keep_tex: true
    theme: metropolis
    latex_engine: pdflatex
    slide_level: 2
    highlight: zenburn
    incremental: false
classoption: "handout"
header-includes:
  \setbeamercolor{frametitle}{bg=purple}
  \hypersetup{colorlinks,citecolor=orange,filecolor=red,linkcolor=brown,urlcolor=blue}
---


## Today's agenda

* **Review: prediction with GLM
  - Blog tip: remember to include prediction intervals! We care about uncertainty in our predictions

* **Prediction: why use demographic variables?**
    1. to simulate \textcolor{purple}{surges} from a particular bloc
    2. to capture correlations between districts (i.e. build a \textcolor{purple}{pooled model})
    3. think about the relationship between shocks and surges
      - Mobilization vs. persuasion 

* **Discussion: do shocks \underline{really} affect election outcomes?**
    * What can we learn from the Dobbs Supreme Court decision (current)?
    * Lessons learned from shark attacks (historical)

* **Discussion: if shocks have effects, does that mean voters are irrational and incompetent?**

* **Discussion: Shocks blog brainstorm**
    * Discussion in small groups


# Prediction: why use demographic variables?

## Why use demographic variables?

Certain demographic blocs have either \underline{historically} (i.e. predictably) and/or 
\underline{contextually} (i.e. unpredictably) mobilized in elections:

* \textbf{African-Americans} historically vote "steadfastly" for Democrats
* \textbf{Working class whites} surged for Trump in 2016
* \textbf{Religious whites} surged for Reagan in 1980
* \textbf{Latinos} historically vote for Democrats, but especially surged during 2018 midterms
* \textbf{... what groups, if any, do we expect to be mobilized, this election cycle? And what events may be mobilizing them (think back to lecture)} 

Election observers tend to call these deviations that can't be predicted 
from past behavior \textcolor{purple}{surges}. 

\textit{A linear regression model specifically estimates coefficients 
for variables to best predict historical (in-sample) behavior,} \newline \textit{but 
we can then perturb those coefficients to simulate surges.}

## Why use demographic variables: surges [1/2]

\textbf{Fitting a predictive model of elections using changes in demographic blocs.} 

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(tidyverse)
library(stargazer)

demog_df <- read_csv("demographic data/demographic_2009_2020.csv")

demog_df_2020 <- read_csv("demographic data/demographic_2020.csv")

demog_df_2009_2019 <- read_csv("demographic data/demographic_2009_2019.csv")

```

```{r, eval=TRUE, echo = FALSE, message=FALSE, warning=FALSE}
incumb_dist_1948_2020_3_ <- read_csv("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/fundamentals II_ incumbency/incumb_dist_1948-2020 (3).csv")
pvdistrict_df <- incumb_dist_1948_2020_3_

# train data
# years <- c(2010, 2012, 2014, 2016, 2018)

dat <- pvdistrict_df %>% 
  #filter(year %in% years) %>% 
    inner_join(demog_df,
              by = c("year" ,"st_cd_fips", "state")) %>%
  # change age var name
   dplyr::rename("age20_29" = "20_29", 
                "age30_44" = "30_44",
               "age45_64" = "45_64", 
                "age65" = "65+",
               "hispanic" = "hispanic or latino",
               "indigenous" = "native american")


# dat$region <- state.division[match(dat$state, state.abb)]
# demog$region <- state.division[match(demog$state, state.abb)]

dat_change <- dat %>%
    group_by(st_cd_fips) %>%
    mutate(Asian_change = asian - lag(asian, order_by = year),
           Black_change = black - lag(black, order_by = year),
           Hispanic_change = hispanic - lag(hispanic, order_by = year),
           Indigenous_change = indigenous - lag(indigenous, order_by = year),
           White_change = white - lag(white, order_by = year),
           Female_change = female - lag(female, order_by = year),
           Male_change = male - lag(male, order_by = year),
           age20_change = age20_29 - lag(age20_29, order_by = year),
           age3045_change = age30_44 - lag(age30_44, order_by = year),
           age4565_change = age45_64 - lag(age45_64, order_by = year),
           age65_change = age65+ - lag(age65, order_by = year)
           )

# # test data
# dat20 <- pvdistrict_df %>% 
#     inner_join(demog_df,
#               by = c("year" ,"st_cd_fips", "state")) %>%
#   # change age var name
#    dplyr::rename("age20_29" = "20_29", 
#                 "age30_44" = "30_44",
#                "age45_64" = "45_64", 
#                 "age65" = "65+",
#                "hispanic" = "hispanic or latino",
#                "indigenous" = "native american")
# names(dat20)
# 
# dat_change20 <- dat20 %>%
#     group_by(st_cd_fips) %>%
#     mutate(Asian_change = asian - lag(asian, order_by = year),
#            Black_change = black - lag(black, order_by = year),
#            Hispanic_change = hispanic - lag(hispanic, order_by = year),
#            Indigenous_change = indigenous - lag(indigenous, order_by = year),
#            White_change = white - lag(white, order_by = year),
#            Female_change = female - lag(female, order_by = year),
#            Male_change = male - lag(male, order_by = year),
#            age20_change = age20_29 - lag(age20_29, order_by = year),
#            age3045_change = age30_44 - lag(age30_44, order_by = year),
#            age4565_change = age45_64 - lag(age45_64, order_by = year),
#            age65_change = age65+ - lag(age65, order_by = year)
#            )
# 
# # subset 2020 for testing
# dat_change20_test <- dat_change20 %>%
#   filter(year == 2020)


demog_df <- demog_df  %>% 
  # change age var name
   dplyr::rename("age20_29" = "20_29", 
                "age30_44" = "30_44",
               "age45_64" = "45_64", 
                "age65" = "65+",
               "hispanic" = "hispanic or latino",
               "indigenous" = "native american")

demog_2022 <- subset(demog_df, year == 2020)
demog_2022 <- as.data.frame(demog_2022)

demog_2022_change <- demog_df %>%
    filter(year %in% c(2018, 2020)) %>%
    group_by(st_cd_fips) %>%
    mutate(Asian_change = asian - lag(asian, order_by = year),
           Black_change = black - lag(black, order_by = year),
           Hispanic_change = hispanic - lag(hispanic, order_by = year),
           Indigenous_change = indigenous - lag(indigenous, order_by = year),
           White_change = white - lag(white, order_by = year),
           Female_change = female - lag(female, order_by = year),
           Male_change = male - lag(male, order_by = year),
           age20_change = age20_29 - lag(age20_29, order_by = year),
           age3045_change = age30_44 - lag(age30_44, order_by = year),
           age4565_change = age45_64 - lag(age45_64, order_by = year),
           age65_change = age65+ - lag(age65, order_by = year)
           ) %>%
    filter(year == 2020)

demog_2022_change <- as.data.frame(demog_2022_change)
```


\scriptsize
```{r}
mod_demog_change <- lm(DemVotesMajorPercent ~ Black_change + Hispanic_change + Asian_change +
                                Female_change +
                                age3045_change + age4565_change + age65_change
                                , data = dat_change)
```

\scriptsize
```{r, results = "asis", echo=FALSE, warning=FALSE, message=FALSE}
stargazer(mod_demog_change, header=FALSE, type='latex', no.space = TRUE,
          column.sep.width = "3pt", font.size = "scriptsize", single.row = TRUE,
          keep = c(1:7, 62:66), omit.table.layout = "sn",
          title = "The electoral effects of demographic change (across districts)")
```

<!-- 0.5*predict(mod_demog, newdata = demog_2020) +
    0.5*predict(mod_poll, newdata = polls_2020) -->
    
## Why use demographic variables: surges [1/2]

**Now, how would our forecast change if there's a Latino surge for Democrats in 2022?** \textbf{Suppose the effect of Latino bloc growth on Dem vote is \underline{doubled} this year (across districts).}

* Historically estimated coefficient for \% Hispanic change: \textbf{2.9}

```{r, eval=FALSE}
predict(mod_demog_change, newdata = demog_2022_change)
```

* \textcolor{purple}{Hypothetical 2022 coefficient} for \% Hispanic change (2x surge): \textcolor{purple}{\textbf{5.8}} \pause


```{r, eval=FALSE}
predict(mod_demog_change, newdata = demog_2022_change) + 
  (5.8-2.9)*demog_2022$hispanic

```

## Why use demographic variables: surges [1/2]

```{r, eval = TRUE, echo=FALSE, message=FALSE, warning=FALSE}
# lm with just demographics, no change
mod_demog <- lm(DemVotesMajorPercent ~ black + hispanic + asian + indigenous +
                         female + age30_44 + age45_64 + age65, data = dat)

his_original <- data.frame(pred = predict(mod_demog, newdata = demog_2022),
    st_cd_fips = demog_2022$st_cd_fips, state = demog_2022$state, district = demog_2022$district)

his1 <- data.frame(pred = predict(mod_demog, newdata = demog_2022) +
    (5.8-2.9)*demog_2022$hispanic,
    st_cd_fips = demog_2022$st_cd_fips, state = demog_2022$state, district = demog_2022$district)

# create district level maps
### with original UCLA shapefiles ###
require(tidyverse)
require(ggplot2)
require(sf)
library(cowplot)

# load geographic data
get_congress_map <- function(cong=116) {
  tmp_file <- tempfile()
  tmp_dir  <- tempdir()
  zp <- sprintf("https://cdmaps.polisci.ucla.edu/shp/districts114.zip",cong)
  download.file(zp, tmp_file)
  unzip(zipfile = tmp_file, exdir = tmp_dir)
  fpath <- paste(tmp_dir, sprintf("districtShapes/districts114.shp",cong), sep = "/")
  st_read(fpath)
}

# load 116th congress
cd116 <- get_congress_map(116)

# change class
cd116$DISTRICT <- as.numeric(cd116$DISTRICT)

# mutate vars
cd116 <- mutate(cd116, district = DISTRICT)
cd116 <- mutate(cd116, state = STATENAME)

# merge for hypothetical
cd116$district <- as.character(cd116$district)
cd116 <- cd116 %>% 
  left_join(his_original, by=c("district", "state"))
# merge for historical
cd116_hyp <- cd116 %>% 
  left_join(his1, by=c("district", "state"))

# plot with simplify
districts_simp <- rmapshaper::ms_simplify(cd116, keep = 0.01)
districts_simp_hyp <- rmapshaper::ms_simplify(cd116_hyp, keep = 0.01)


plot_original <- ggplot() + 
  geom_sf(data=districts_simp,aes(fill = (pred >= 50)),
          inherit.aes=FALSE,alpha=0.9) + 
  #scale_fill_gradient(low = "white", high = "black", limits=c(0,90)) +
  coord_sf(xlim = c(-172.27, -66.57), ylim = c(18.55, 71.23), expand = FALSE) +  
  theme_void() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

plot_1 <-ggplot() + 
  geom_sf(data=districts_simp_hyp,aes(fill = (pred.x >= 50)),
          inherit.aes=FALSE,alpha=0.9) + 
  #scale_fill_gradient(low = "white", high = "black", limits=c(0,90)) +
  coord_sf(xlim = c(-172.27, -66.57), ylim = c(18.55, 71.23), expand = FALSE) +  
  theme_void() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

```{r, eval=FALSE, warning=FALSE, message=FALSE}
plot_grid(plot_original, plot_1)

```

![](imgs/plot comp)

\scriptsize
```{r, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
### pull 2020 CD map with tigris ###
library(tigris)
library(sf)
library(ggplot2)
shp_cd <- tigris::congressional_districts(
      state = NULL,
      cb = FALSE,
      resolution = "500k",
      year = 2020)
# check that it plots itself
# shp_cd %>% tigris::shift_geometry() %>% ggplot() + geom_sf()
# store the plot output
shp_cd <- shp_cd %>% tigris::shift_geometry()
# mutate GEOID
shp_cd <- shp_cd %>%
  dplyr::rename(st_cd_fips = GEOID)
# remember to simplify
# shp_cd_simp <- rmapshaper::ms_simplify(shp_cd, keep = 0.01)
# merge cd shape file by fips with original predictions
shp_cd_simp_pred <- his_original %>%
  inner_join(shp_cd, by = "st_cd_fips")
# merge cd shape file by fips with hypothetical predictions
shp_cd_simp_pred_his <- his1 %>%
  inner_join(shp_cd, by = "st_cd_fips")

# create map
 library(RColorBrewer)

 plot_original <- ggplot(data = shp_cd_simp_pred) +
   geom_sf(color = NA, aes(geometry = geometry, fill = (pred >= 50), lwd = 0)) +
  scale_fill_manual(values=brewer.pal(2,'YlOrRd'), na.value="grey") +
   theme(rect = element_blank(),
        panel.grid.major = element_line(colour = "transparent"),
        axis.text.x = element_blank(),
         axis.text.y = element_blank(),
         axis.ticks = element_blank()) +
   labs(title = "2022 Midterm Election Prediction",
      subtitle = "historical Hispanic demographic change effect (2.9)",
       fill = "")

 plot_1 <- ggplot(data = shp_cd_simp_pred_his) +
   geom_sf(color = NA, aes(geometry = geometry, fill = (pred >= 50), lwd = 0)) +
 scale_fill_manual(values=brewer.pal(2,'YlOrRd'), na.value="grey") +
   theme(rect = element_blank(),
         panel.grid.major = element_line(colour = "transparent"),
        axis.text.x = element_blank(),
         axis.text.y = element_blank(),
         axis.ticks = element_blank()) +
   labs(title = "2022 Midterm Election Prediction",
        subtitle = "hypothetical Hispanic demographic change effect (5.8)",
        fill = "")
 
library(cowplot)
plot_grid(plot_original, plot_1)
```


## Why use demographic variables: surges [1/2]

Previously, we fit a separate poll-based regression for \underline{each} district $\rightsquigarrow$
\textit{assumes separate parameters for each district}:

$$
DemPV_{district} = f(\alpha_{district} + \beta_{1,district}x_1 + \ldots + \beta_{k,district} x_{k,district})
$$
\newline With our demographic bloc model, we fit one regression \underline{across} districts $\rightsquigarrow$
\textit{assumes one set of parameters for all districts}:

$$
DemPV_{district} = f(\alpha + \beta_1 x_1 + \ldots + \beta_k x_k)
$$
\textbf{Q:} Why are some advantages of the latter?


## Why \textit{else} use demographic variables?

Previously, we fit a separate poll-based regression for \underline{each} district $\rightsquigarrow$
\textit{assumes separate parameters for each district}\footnote{We lied to you last week! This is technically the proper notation for district-by-district models.} (\textcolor{purple}{unpooled model}):

$$
DemPV_{district} = f(\alpha_{district} + \beta_{1,district}x_1 + \ldots + \beta_{k,district} x_{k,district})
$$
With our demographic bloc model, we fit one regression \underline{across} district $\rightsquigarrow$
\textit{assumes one set of parameters for all districts} (\textcolor{purple}{pooled model}):

$$
DemPV_{district} = f(\alpha + \beta_1 x_1 + \ldots + \beta_k x_k)
$$
\textbf{Q:} Why are some advantages of the latter?
<!--
- one set of parameters for a model means we don't assume districts are totally independent!
- in some districts, we may not have enough data to estimate a model JUST for that district
- if there's an error, its likely to be correlated across districts rather than just totally idiosyncratic
-->

## Why use demographic variables: across-district correlations [2/2]


```{r, eval = TRUE, echo=FALSE, warning=FALSE, message=FALSE}
pvdistrict_df <- incumb_dist_1948_2020_3_

# subset to one state for example
yrs <- c(2008, 2010, 2012, 2014, 2016, 2018, 2020)
bama <- pvdistrict_df %>%
  filter(year %in% yrs) %>%
  filter(state == c("Alabama"))
  
pvcorr_df <- data.frame()
for (st_cd_fips.x in unique(bama$st_cd_fips)) {
  for (st_cd_fips.y in unique(bama$st_cd_fips)) {
    ## to make sure we have complete pairwise set of observations
    yr.x <- bama$year[bama$st_cd_fips == st_cd_fips.x]
    yr.y <- bama$year[bama$st_cd_fips == st_cd_fips.y]
    yrs <- base::intersect(yr.x, yr.y)  ## the `base::` is to avoid conflict with a `dplyr` function also called `intersect`
    r.sq <- cor(bama$DemVotesMajorPercent[bama$st_cd_fips == st_cd_fips.x & bama$year %in% yrs],
                bama$DemVotesMajorPercent[bama$st_cd_fips == st_cd_fips.y & bama$year %in% yrs],
                use = "pairwise.complete.obs")
    pvcorr_df <- bind_rows(pvcorr_df, data.frame(st_cd_fips.x, st_cd_fips.y, r.sq))
  }
}

cd_order <- demog_2022_change$st_cd_fips[order(demog_2022_change$black)]
pvcorr_df$st_cd_fips.x <- factor(pvcorr_df$st_cd_fips.x, levels = cd_order)
pvcorr_df$st_cd_fips.y <- factor(pvcorr_df$st_cd_fips.y, levels = cd_order)

pvcorr_df %>% 
  filter(!is.na(st_cd_fips.x), !is.na(st_cd_fips.y)) %>%
  ggplot(aes(x = st_cd_fips.x, y = st_cd_fips.y, fill = r.sq)) + 
  geom_tile() +
  xlab("") + ylab("") + 
  scale_fill_gradient(low="yellow", high="purple", name="R-squared in\npopular vote\n2008-2020\nAlabama") +
  xlab("<-- least Black districts in 2020          most Black districts in 2020 -->") +
  theme(axis.text.x = element_text(angle=90, hjust=1),
        axis.title.x = element_text(size=16))
```

```{r, eval=TRUE, echo=FALSE}
cd_order <- demog_2022_change$st_cd_fips[order(demog_2022_change$black)]
pvcorr_df$st_cd_fips.x <- factor(pvcorr_df$st_cd_fips.x, levels = cd_order)
pvcorr_df$st_cd_fips.y <- factor(pvcorr_df$st_cd_fips.y, levels = cd_order)

pvcorr_df %>% 
  filter(!is.na(st_cd_fips.x), !is.na(st_cd_fips.y)) %>%
  ggplot(aes(x = st_cd_fips.x, y = st_cd_fips.y, fill = r.sq)) + 
  geom_tile() +
  xlab("") + ylab("") + 
  scale_fill_gradient(low="yellow", high="purple", name="R-squared in\npopular vote\n2008-2020\nAlabama") +
  xlab("<-- least Black districts in 2020          most Black districts in 2020 -->") +
  theme(axis.text.x = element_text(angle=90, hjust=1),
        axis.title.x = element_text(size=16))
```

## Why use demographic variables: across-district correlations [2/2]

```{r, eval=TRUE, echo=FALSE}
cd_order <- demog_2022_change$st_cd_fips[order(demog_2022_change$hispanic)]
pvcorr_df$st_cd_fips.x <- factor(pvcorr_df$st_cd_fips.x, levels = cd_order)
pvcorr_df$st_cd_fips.y <- factor(pvcorr_df$st_cd_fips.y, levels = cd_order)

pvcorr_df %>% 
  filter(!is.na(st_cd_fips.x), !is.na(st_cd_fips.y)) %>%
  ggplot(aes(x = st_cd_fips.x, y = st_cd_fips.y, fill = r.sq)) + 
  geom_tile() +
  xlab("") + ylab("") + 
  scale_fill_gradient(low="yellow", high="purple", name="R-squared in\npopular vote\n2008-2020\nAlabama") +
  xlab("<-- least Hispanic districts in 2020          most Hispanic districts in 2020 -->") +
  theme(axis.text.x = element_text(angle=90, hjust=1),
        axis.title.x = element_text(size=16))
```

## Why use demographic variables: across-state correlations [2/2]

To recap:

* A pooled model captures correlation across districts by using the same $\alpha$ and $\beta_1,\ldots,\beta_k$
to predict each district's outcome.

* A pooled model relies on less data from each district, thus "drawing strength" on less data-sparse district

* \textbf{You can think about this as the "data borrowing" concept we've talked about in past weeks - you can use these correlations to update district's A's prediction \textit{if} district B's outcome is hypothetically known} 
    * ex: \texttt{ricardofernholz.com/election} - an example to read through
    
\small 

\textbf{Q:} Can I combine a \textit{pooled (i.e. correlated) model} with \textit{unpooled (i.e. idiosyncratic) models} fit to each district?

\textbf{A:} Yes, through the power of ensembling! \textcolor{purple}{And with appropriate checks on the model performance!}

\tiny

```{r, eval=FALSE}
for (s in st_cd_fips) {
  DemPV_s <- 0.5 * predict(mod_demog_change, demog_2020_change %>% filter(st_cd_fips==s)) + 
             0.5 * predict(mod_state_poll_glms[s], poll_pvstate_vep_df %>% filter(st_cd_fips==s, party=="democrat"))
}
```

This is toy/example code.

# Discussion: Do shocks really affect election outcomes?

## Shocks?

**Q:** What are \textit{examples} of **apolitical shocks** that would affect elections? \pause

* **Natural disaster**: Shark attack (Achen and Bartels 2017), tornado (Healy and Malhotra 2010) $\rightsquigarrow$ \textcolor{red}{decreased} support for incumbents \pause
* **Sports** (Healy, Mo, Malhotra 2010): college football team losing their game $\rightsquigarrow$  \textcolor{red}{decreased} support for incumbents \pause
* **Lottery** (Bagues and Esteve-Volart 2016): Towns winning Spanish Christmas lottery $\rightsquigarrow$  \textcolor{green}{increased} support for incumbents \pause

**Q: ** Why do we care about apolitical shocks? 

<!-- It seems to be the evidence of **voter irrationality.** -->


**Q:** What are \textit{examples} of **political shocks** that would affect elections? \pause

* **Natural disaster**: Shark attack (Achen and Bartels 2017), tornado (Healy and Malhotra 2010) $\rightsquigarrow$ \textcolor{red}{decreased} support for incumbents \pause
* **Sports** (Healy, Mo, Malhotra 2010): college football team losing their game $\rightsquigarrow$  \textcolor{red}{decreased} support for incumbents \pause
* **Lottery** (Bagues and Esteve-Volart 2016): Towns winning Spanish Christmas lottery $\rightsquigarrow$  \textcolor{green}{increased} support for incumbents \pause

**Q: ** What political shocks have there been during the current election cycle? 

**Q: ** Why do we care about political shocks? 

## The Dobbs Supreme Court decision (Roe v. Wade)

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(dotenv)
library(jsonlite)
library(tidyverse)
library(lubridate)
# # load up hidden api key
# article_api <- "zbNGtPzlJflHASLOR1jGJDAngZjfCuNK"
# #semantic_api <- Sys.getenv("SEMANTIC_API")
# 
# # set base url
# base_url_art <- "http://api.nytimes.com/svc/search/v2/articlesearch.json?fq="
# #base_url_sem <- "http://api.nytimes.com/svc/semantic/v2/concept/name"w
# 
# # set parameters
# term <- "dobbs"
# facet_field <- "day_of_week"
# facet <- "true"
# begin_date <- "20220101"
# end_date <- "20221015"
# complete_url <- "https://api.nytimes.com/svc/search/v2/articlesearch.json?fq=dobbs&facet_field=day_of_week&facet=true&begin_date=20220101&end_date=20221015&api-key=zbNGtPzlJflHASLOR1jGJDAngZjfCuNK"
# 
# 
# complete_url2 <-paste0(base_url_art,fq =term,"&facet_field=",facet_field,"&facet=",facet,"&begin_date=",begin_date,"&end_date=",end_date,"&api-key=",article_api,sep = "")
# 
# # import dataset to R
# sus <- fromJSON(complete_url2) 
# 
# # view how many hits
# sus$response$meta$hits
# 
# hits <- sus$response$meta$hits
# cat("There were ",hits," hits for the search term Dobbs during 2022 to date",sep = "")
# 
# max_pages <- round((hits / 10) - 1)
# 
# # store all pages in list
# pages <- list()
# for(i in 0:max_pages){
#     sus_df <- fromJSON(paste0(complete_url2, "&page=", i),
#     flatten = TRUE) %>% 
#     data.frame() 
#   message("Retrieving page ", i)
#   pages[[i+1]] <- sus_df
#   Sys.sleep(6)
# }
# 
# 
# # trying again - WORKS!!!
# sus0 <- fromJSON(paste0(complete_url2, "&page=0"), flatten = TRUE)
# nrow(sus0$response$docs)
# sus1 <- fromJSON(paste0(complete_url2, "&page=1"), flatten = TRUE)
# nrow(sus1$response$docs)
# sus2 <- fromJSON(paste0(complete_url2, "&page=2"), flatten = TRUE)
# nrow(sus2$response$docs)
# 
# organizations <- rbind_pages(
#   list(sus0$response$docs, sus1$response$docs, sus2$response$docs)
# )
# nrow(organizations)
# 
# pages <- list()
# Sys.sleep(1) 
# for(i in 0:24){
#   mydata <- fromJSON(paste0(complete_url2, "&page=", i))
#   message("Retrieving page ", i)
#   pages[[i+1]] <- mydata$response$docs
#   Sys.sleep(6) 
# }
# 
# #combine all into one
# organizations <- rbind_pages(pages)
# 
# #check output
# nrow(organizations)
# 
# colnames(organizations)
# 
# 
# # trying with hits
# sus0 <- fromJSON(paste0(complete_url2, "&page=0"), flatten = TRUE)
# nrow(sus0$response)
# sus1 <- fromJSON(paste0(complete_url2, "&page=1"), flatten = TRUE)
# nrow(sus1$response$docs)
# sus2 <- fromJSON(paste0(complete_url2, "&page=2"), flatten = TRUE)
# nrow(sus2$response$docs)
# 
# organizations <- rbind_pages(
#   list(sus0$response, sus1$response, sus2$response)
# )
# nrow(organizations)
# 
# pages <- list()
# Sys.sleep(1) 
# for(i in 0:24){
#   mydata <- fromJSON(paste0(complete_url2, "&page=", i)) 
#   message("Retrieving page ", i)
#   pages[[i+1]] <- mydata$response$docs
#   Sys.sleep(6) 
# }
# 
# pages <- as.data.frame(pages)
# do.call(rbind.data.frame, pages)
# library (plyr)
# pages <- ldply(pages, data.frame)
# data.frame(t(sapply(pages,c)))
# rbind.fill(pages)
# 
# 
# #combine all into one
# mydata <- rbind_pages(pages)
# 
# #check output
# nrow(mydata)
# 
# # save df
# saveRDS(mydata, file = "dobbs_2022.RDS")

# reload
mydata <- readRDS("dobbs_2022.RDS")

# check colnames
# colnames(mydata)

# visualization by month
library(dplyr)
month <- mydata %>% 
  group_by(month = month(pub_date, label = T)) %>% 
  dplyr::summarize(count = n()) %>% 
  ggplot(aes(month, count, group = 1, color = count)) +
    geom_line() +
    labs(y = "Article Count", x = "",
         title = "NYT Articles mentioning Dobbs Supreme Court decision in 2022",
         color = "")

# visualization by day
day <- mydata %>% 
  group_by(month_day = paste0(month(pub_date, label = T),
           day = day(pub_date))) %>% 
  dplyr::summarize(count = n()) %>% 
  ggplot(aes(month_day, count, group = 1, color = count)) +
    geom_line() +
    labs(y = "Article Count", x = "",
         title = "NYT Articles mentioning Dobbs Supreme Court decision in 2022",
         color = "")

# how about visualization by week
# extract raw date
mydata <- mydata %>% 
  mutate(publ_date = substr(pub_date, 1, 10))

# mutate week variable
mydata <- mydata %>% 
  mutate(week = strftime(publ_date, format = "%V"))

# plot
week <- mydata %>% 
  group_by(week) %>% 
  dplyr::summarize(count = n()) %>% 
  ggplot(aes(week, count, group = 1, color = count)) +
    geom_line() +
    labs(y = "Article Count", x = "Week",
         title = "NYT Articles mentioning Dobbs Supreme Court decision in 2022",
         color = "") + # now add line for when decision was leaked
      geom_segment(x=("18"), xend=("18"),y=0,yend=37, lty=2, color="purple", alpha=0.4) +
      annotate("text", x=("18"), y=35, label="Decision leaked", size=3) +
  geom_segment(x=("25"), xend=("25"),y=0,yend=37, lty=2, color="red", alpha=0.4) +
      annotate("text", x=("25"), y=35, label="Decision released", size=3) # now add line for when decision was actually made

```
```{r, eval=TRUE, echo=FALSE}
# plot
mydata %>% 
  group_by(week) %>% 
  dplyr::summarize(count = n()) %>% 
  ggplot(aes(week, count, group = 1, color = count)) +
    geom_line() +
    labs(y = "Article Count", x = "Week",
         title = "NYT Articles mentioning Dobbs Supreme Court decision in 2022",
         color = "") + # now add line for when decision was leaked
      geom_segment(x=("18"), xend=("18"),y=0,yend=37, lty=2, color="purple", alpha=0.4) +
      annotate("text", x=("18"), y=35, label="Decision leaked", size=3) +
  geom_segment(x=("25"), xend=("25"),y=0,yend=37, lty=2, color="red", alpha=0.4) +
      annotate("text", x=("25"), y=35, label="Decision released", size=3) # now add line for when decision was actually made
```


## Questions
\tiny
**Q: ** What do you notice about the number of articles published before and after the Dobbs decision was leaked and released? \newline

**Q: ** Are these "shocks," in terms of article count, long-lasting? What seems like a reasonable metric for shock longevity? \newline

**Q: ** What can we infer, if anything, about the duration and strength of this particular shock, especially with regard to voter opinion/behavior? \newline

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
# now compare this to generic ballot
X538_generic_ballot_averages_2018_2022 <- read_csv("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/Polling/Section data/final files/538_generic_ballot_averages_2018-2022.csv")
gb <- X538_generic_ballot_averages_2018_2022

# convert dat
gb <- gb %>%
  mutate(date_ = mdy(date)) %>%
  mutate(year = substr(date_, 1, 4)) %>%
  filter(year == 2022) %>%
  mutate(week = strftime(date_, format = "%V")) # Jan 1 looks weird 

# #get avg by party and week
# dem <- gb %>%
#   filter(candidate == 'Democrats')
# x <- plyr::ddply(dem, .(week), function(z) mean(z$pct_estimate))
# x$candidate <- c('Democrats')
# x$avg_dem <- x$V1
# x <- x %>%
#   select(-V1)
# x$avg_dem <-  round(x$avg_dem , digits = 1)
#
#
# rep <- gb %>%
#   filter(candidate == 'Republicans')
# y <- plyr::ddply(rep, .(week), function(z) mean(z$pct_estimate))
# y$candidate <- c('Republicans')
# y$avg_rep <- y$V1
# y <- y %>%
#   select(-V1)
# y$avg_rep <-  round(y$avg_rep, digits = 1)
#
# #put all data frames into list
# df_list <- list(gb, x, y)
#
# #merge all data frames together
# polls_df <- df_list %>% reduce(full_join, by=c("candidate", "week"))
#
# # remove NAs
# polls_df[] <-  t(apply(polls_df, 1, function(x) c(x[!is.na(x)], x[is.na(x)])))
#
# polls_df <- polls_df %>%
#   select(-avg_rep)
#
# polls_df$avg_support <- polls_df$avg_dem
#
# polls_df <- polls_df %>%
#   select(-avg_dem)
#
# # keep only unique dates
# polls_df <- polls_df %>%
#   distinct(cycle, week, date_, avg_support, candidate) %>%
#   filter(week != 52)

# visualize polls
# polls_df %>%
#   #group_by(candidate == 'Democrats') %>%
#   #mutate(date_ = as.Date(date_)) %>%
#   ggplot(aes(x = week, y = avg_support,
#              colour = candidate)) +
#   geom_line(aes(group=candidate), size = 0.3) + geom_point(size = 0.3) +
#     #scale_x_date(date_labels = "%b, %Y") +
#   ylab("generic ballot support") + xlab("") +
#     theme_classic() +
#   # now add line for when decision was leaked and released
#       geom_segment(x=("18"), xend=("18"),y=0,yend=33, lty=2, color="purple", alpha=0.4) +
#       annotate("text", x=("18"), y=31, label="Decision leaked", size=2) +
#   geom_segment(x=("25"), xend=("25"),y=0,yend=33, lty=2, color="red", alpha=0.4) +
#       annotate("text", x=("25"), y=31, label="Decision released", size=2)

```

![](imgs/generic ballot dobbs)

## Questions pt.2

**Q: ** What changes do we see in the generic ballot in response to the Dobbs decision? \newline
**Q: ** What additional evidence would be helpful in understanding Dobbs' effect on voter behavior? \newline
**Q: ** Do you think we will see a \textbf{surge} in turnout or registration amongst certain demographic groups this cycle? \newline
![](imgs/roe)
<!-- {width=750px, height=550px} -->

![](imgs/politico)
<!-- {width=750px, height=550px} -->


## A canonical debate: shark attacks (NJ, 1916)

\footnotesize
* **Achen and Bartels** say: \textcolor{red}{Beach towns} in NJ affected by shark attacks \textit{voted significantly less for the incumbent} than non-beach towns 
* **Fowler and Hall** respond: 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readstata13)

### override default mgp in par to make better ticks
par.gh <- function(..., mgp=c(0,.05,.05)) {
  par(..., mgp=mgp)
}

### function to produce nice plots
### preserves most defaults through "..."
### defaults labels to blank, axes to missing, makes points look nice
plot.gh <- function(..., xlab="", ylab="", xaxt="n", yaxt="n", pch=21, bg="gray80", col="black",bor_col="white", x_seq = NULL,  y_seq = NULL, bty=NULL, rect_col="gray90", do_axes=NULL, do_seq=T, outer.box=F, outer.box.col="black", grid_col="white", point.size=2, remove.axis=F) {
  plot(..., xlab=xlab, ylab=ylab, xaxt=xaxt, yaxt=yaxt, pch=pch, bg=bg,  bty="n")
  rect(par("usr")[1], par("usr")[3], par("usr")[2], par("usr")[4], col = rect_col, border= bor_col)
  # Draw sequences in background
  if(do_seq!=F){
    grid(col=grid_col, lty=2)
  }
  # Add axis
  if (!remove.axis) {
    axis(side=1,lwd=0, lwd.ticks=1, tck=-0.01, col="white", col.axis="gray20", cex.axis=.8)
    axis(side=2,lwd=0, lwd.ticks=1, tck=-0.01, las=1, col="white", col.axis="gray20", cex.axis=.8)
  }
  # Add points
  points(...,pch=21, bg=bg, col=col, lwd=1.5, cex=point.size)
  # Y/N add box around plot
  if(outer.box==T){
    box(col= outer.box.col)
  }
}

### Michael Gill's brilliant function to make transparent colors
makeTransparent<-function(someColor, alpha)
{
  newColor<-col2rgb(someColor)
  apply(newColor, 2, function(curcoldata){rgb(red=curcoldata[1], green=curcoldata[2],
                                              blue=curcoldata[3],alpha=alpha, maxColorValue=255)})
}
```


```{r, echo=FALSE, fig.height=5}
### town-level plot
data <- read.dta13("NJ_TownData_RGraph.dta")
data$mat <- 0
data$mat[grep("Matawan", data$town)] <- 1
data.o <- data[data$county=="Ocean",]
cexs.o <- 1.2*(data.o$total1916 / max(data.o$total1916)) + .5
cols.o <- ifelse(data.o$beach==1, "red", "gray50")
cols.o[data.o$nearbeach==1] <- "royalblue"
pchs.o <- ifelse(data.o$beach==1, 15,16)
pchs.o[data.o$nearbeach==1] <- 17
par.gh(mfrow=c(1,2)) # , oma=c(3,3,3,3), mar=c(1,1,1,1)
plot.gh(x=data.o$vote1912, y=data.o$vote1916,bg="gray90", xlim=c(.1,.8), ylim=c(.1,.7), col="gray90")
points(x=data.o$vote1912, y=data.o$vote1916, col=cols.o, pch=pchs.o, cex=cexs.o)
points(x=data.o$vote1912[data.o$town=="Sea Side Park"], y=data.o$vote1916[data.o$town=="Sea Side Park"], cex=2)
legend("topleft", bty="n", pch=c(15,17,16), col=c("red", "dodgerblue", "gray50"), c("Beach Towns", "Near-Beach Towns", "Non-Beach Towns"), cex=.8)
arrows(x0=.36, x1=.41, y0=.54, y1=.56, length=.1, angle=20)
mtext(side=2, line=2, "Wilson 2-Party Vote Share, 1916", cex=.9)
mtext(side=1, line=1, "Wilson 3-Party Vote Share, 1912")
mtext(side=3, line=.1, "Ocean County, NJ", font=2)
text(x=.27, y=.54, "Sea Side Park", cex=.7)
cexs <- 1.2*(data$total1916 / max(data$total1916)) + .5
pchs <- ifelse(data$beach==1, 15,16)
pchs[data$nearbeach==1] <- 17
cols <- ifelse(data$beach==1, "red", "gray50")
cols[data$nearbeach==1] <- "royalblue"
plot.gh(x=data$vote1912, y=data$vote1916,bg="gray90", xlim=c(.1,.8), ylim=c(.1,.7), remove.axis=T, col="gray90")
points(x=data$vote1912, y=data$vote1916, col=cols, pch=pchs, cex=cexs)
points(x=data[data$mat==1,]$vote1912, y=data[data$mat==1,]$vote1916, cex=2)
axis(side=1, lwd=0, cex.axis=.8)
mtext(side=1, line=1, "Wilson 3-Party Vote Share, 1912")
mtext(side=3, "All NJ Counties", font=2)
legend("topleft", bty="n", pch=c(15,17,16,1), col=c("red", "dodgerblue", "gray50", "black"), c("Beach Towns", "Near-Beach Towns", "Non-Beach Towns", "Matawan Attacks"), pt.cex=c(1,1,1,2), cex=.8)
```
<!-- Lessons learned:
* to look at how places vote, look at how they voted last year (stickiness)
* check whether your data is/has outliers; also check how sensitive your data is to exclusion of outliers
* Twyman's Law: if it's a buzzworthy or 'exciting finding' its almost always wrong
-->

* \footnotesize Including \textcolor{red}{Sea Side Park} (that was omitted) dampens the effect
* \footnotesize Including all counties shows that (1) the \textcolor{red}{two beach counties} were very unusual outliers (2) \textcolor{red}{beach towns and near-beach/non-beach towns voted similarly} 

# Discussion: If shocks have effects, does that mean voters are irrational and incompetent?

## Rational?

**Q:** How do voters \textit{rationally} consider the impact of shocks in their vote? \pause

* **Consider what gov't could have done better** (Many seemingly apolitical shocks are not apolitical): \pause

    * Better \textcolor{purple}{response} (Healy and Malhotra 2010): effects depend on whether the disaster declaration took place  \pause
    * Better \textcolor{purple}{relief} (Gasper and Reeves 2011): when presidents reject governors' requests for federal aid $\rightsquigarrow$ punish presidents and reward governors
    * Better \textcolor{purple}{preparation} 

## Rational?

**Q:** How do voters \textit{rationally} consider the impact of shocks in their vote? 

* **Global comparison** (Powell \& Whitten 1993): Punish \textcolor{purple}{only when} the gov't is doing \textcolor{purple}{worse than other gov'ts}
    * Ex: the speed of recovery from global economic crisis compared to other countries  \pause
* **Forward-looking voting** (Leininger and Schaub 2020): Germans \textcolor{purple}{vote to align local incumbents with higher levels of government}—with expectation that this will help them through the COVID crisis 

## Irrational?

**Q:** How do voters \textit{irrationally} consider the impact of shocks in their vote? 

* \textbf{Partisan bias} (Malhotra and Kuo 2008; Healy, Kuo, Malhotra 2014).

    * Survey question: \texttt{"How much is [Official X] \textbf{to blame} for making American vulnerable to \textbf{the attacks on September 11?}"}
    * Experimental variation: Randomly change the partisanship of \texttt{[Offical X]}
    * Finding: The answer depends on the partisanship

* \textbf{Myopic voting}: do not respond to policies related to disaster \textcolor{red}{preparedness}.

    * Respond only to policies and relief right after the crisis has already happened (Healy and Malhotra 2010) $\rightsquigarrow$ create \textcolor{red}{perverse incentives} to (i) invest less on (more effective + less expensive) preventive measures and (ii) claim credit on performative ex-post measures


## Breakout room: Shocks blog brainstorm


\small
\underline{Small group discussion:} How do you think shocks will factor into the election this year? Do you expect shocks to matter more for certain (groups of) voters or in certain races?

## Blog Extensions

* **NYT scrape and anticipated shocks**: Replicate the Dobbs NYT example from section, but with a different topic/issue that you believe has been a "shock" during the current cycle. Do your findings alter your predictions (hypothetically)? Explain why/ why not. 

To replicated the section example, you will need to register for a NYT developer account and acquire an API key.

See additional documentation on replicating the section example: 
https://developer.nytimes.com/get-started \newline
https://developer.nytimes.com/docs/articlesearch-product/1/overview \newline
https://rpubs.com/justinmorgan/nytimesarticleapi


* **Incorporate a pooled model into your prediction this week** If you want to just continue refining your current model, that is also fine, but write up a brief explanation as to why you've chosen not to incorporate a pooled model.

