---
title: "Fundamentals II: Incumbency"
subtitle: "Gov 1347: Election Analysis"
author: Kiara Hernandez
date: \today
institute: Harvard University
fontsize: 10pt
output:
 beamer_presentation:
    keep_tex: true
    theme: metropolis
    latex_engine: pdflatex
    slide_level: 2
    highlight: zenburn
    incremental: false
header-includes:
  \setbeamercolor{frametitle}{bg=purple}
  \hypersetup{colorlinks,citecolor=orange,filecolor=red,linkcolor=brown,urlcolor=blue}
# subtitle: 'Gov 1347: Election Analytics'
---
<!---classoption: "handout"-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse); library(knitr)
```

## Today's agenda

* \textbf{Review ensemble models}

* \textbf{(Slowly) moving from prediction with linear models to prediction with probabilistic models}

    -Discuss Blog 03, Extension #1: differences between FiveThirtyEight and The Economist forecasting models
    -Build out toy preliminary versions of FiveThirtyEight, The Economist and Klarner et al. (2006) models

* \textbf{(Slowly) moving from predicting voteshare and seatshare to predicting outcomes for individual seats}
    
    -Case 1: A district with good polling data
    -Case 2: A district with bad polling data
    -Aggregating individual seat predictions

* \textbf{Incumbency advantage and expert predictions}
    -Relationship between incumbent voteshare (seatshare) and different expert predictions
    
* \textbf{Preview of next week: Probabilistic models}

## Ensemble models

* Excellent job on blog posts! \pause

* Review: \textbf{weighted ensembles} are combinations of models (\underline{any} kind) where you pick the weights (in some meaningful way)!
    - \scriptsize a model could be the \underline{raw poll}
    - \scriptsize a multivariate regression is not an ensemble of \textit{models}, though you can interpret coefficients as weights
    - \scriptsize \textbf{sensitivity analyses} are important: is your prediction heavily reliant on one model or a particular set of weights?

## Previously on Gov 1347...

\normalsize \textbf{Extension 2/3}: 4 options for a "weighted ensemble" of individual polls (with $X$ weeks left)
\scriptsize

\textbf{1 (2022 polls adjusted by recent expert grades):} 

$PV_{2022}^{(inc)} = \Big(\sum_{i=1}^{n} w_i \times Poll^{(inc)}_{2022,i} /  \sum_{i=1}{w_i}\Big)$ where 
    $$
    w_i = \Bigg\{ \begin{array}{ll}
      0.75 & \text{if } Grade2018_{i} = A \\
      0.2 & \text{if } Grade2018_{i} = B\\
      0.05 & \text{if } Grade2018_{i} \leq C \\
      \end{array}
    $$
\pause

\textbf{2 (2022 polls adjusted by recent predictiveness):} 

(same as above but $w_i \approx$ poll error or sample size)

\pause

\textbf{3} 

$PV_{2022}^{(inc)} = \sum_{i=1}^{N} w_i \times \underbrace{(\widehat{\alpha} + \widehat{\beta}}_{estimated \ from \ PV^{(inc)}_{t} = \alpha + \beta \cdot AvgPoll_{t}^{(inc)}} \times Poll_{2022,i}^{(inc)})$ <!-- Q: what does this mean? A: it means we're adjusting not only for recent poll quality but also predictiveness/over and underperformance of polls based on historical data -->

\pause

$\rightsquigarrow$ \textbf{(2022 polls adjusted by recent grades and average historical underperformance)}

\pause

\textbf{4 (2022 polls adjusted by two most recent grades):}
$$
\begin{aligned}
PV_{2022}^{(inc)} = & 0.75 \times \Big(\sum_{i=1}^{n} w_i^{(2018)} Poll^{(inc)}_{2020,i} /  \sum_{i=1}{w_i}^{(2018)}\Big) + 
& 0.25 \times \Big(\sum_{i=1}^{n} w_i^{(2014)} Poll^{(inc)}_{2016,i} /  \sum_{i=1}{w_i}^{(2014)}\Big)
\end{aligned}
$$

# Discussion (15 minutes)

In pairs...

1. Between the [FiveThirtyEight]() and [The Economist]() election forecasts,  which one seemed better to you? Any better ones out there?
2. Share something you found (a) \underline{interesting} and something you found (b) \underline{confusing} while updating your blogs last week.
3. Review Klarner and Buchanan (2006) from discussion this week (in Lab sessions folder).

# How does FiveThirtyEight's model take polling and incumbency into account?
\tiny
Polling

-Step 1: Collect, weight and average polls.
  -weight based on sample size
  -weight based on recency
  -pollster rating
  
-Step 2: Adjust polls.

-Step 3: Combine polls with demographic and (in the case of polls-plus) economic data.

-Step 4: Account for uncertainty and simulate the election thousands of times.

"Borrowing polls" - partisan lean metric + adjustment with district similarity score
(we'll see this in a few slides)

Incumbency

-Factors into "fundamentals"

  -Incumbent's lagged margin of victory
  
  -Congressional approval ratings (attitudes toward incumbent)
  
  -Scandals
  
  -Roll call voting record (in-line voting w/ party)

# How does The Economist's model take polling and incumbency into account?

Polling
  
  Step 1: Estimate national and district-level trends in support for candidates
    (1) Congressional generic ballot
    (2) Weighting! By proximity to election day, changes in national political environment (party generic ballot), pollster quality (record of over/underestimating)
  
Incumbency
  
  ???


# How does Klarner and Buchanan's model work in practice?
\tiny
Two approaches outlined:

(1) District-level approach: Polls

(2) Aggregate approach: "National partisan tide" (similar to a fundamentals-only model)
    -health of the economy, presidential approval, quality of candidates, voting intentions
    
Their approach: combine both (similar to what we did last week!)
  -DV: Democrat two-party voteshare
  -IVs: (a) district partisan composition
          (i) past House vote: Democrat two-party voteshare from most recent House election
          (ii) past presidential vote: district presidential voteshare - national Democratic two-party voteshare 
  
  (b) candidate attributes
    (i) Incumbency
    (ii) quality candidate, closed: candidate facing an incumbent
    (iii) quality candidate, open: two non-incumbents
    (iv) past House member, closed
    (v) past House member, open
  
  (c) national partisan tides
    (i) Democratic vote intention: % of respondents who expressed intention to vote for Democratic House candidate (Gallup poll ~March 10)
    (ii) Presidential approval: coded towards Democrats and conditional on party of sitting president (if Dem president, approval rate; if Rep president, disapproval rate) (Gallup poll ~March 10)
    (iii) change in RDI: percent change per capita RDI, February of year before election year - February of election year
    (iv) Midterm penalty: president's party should lose votes

# Individual seat prediction: A district with (pretty) good polling data
## Example: Ohio District 01 (3901)

```{r, eval = TRUE, echo=FALSE, message=FALSE, warning=FALSE}
# load polling data
dist_polls_2018_2022 <- read_csv("dist_polls_2018-2022.csv")

# subset
polls_df <- dist_polls_2018_2022 %>%
  select(pollster, sponsors, display_name, fte_grade,
         start_date, end_date, sample_size, population, cycle, election_date,
         party, candidate_name, pct, st_cd_fips) %>%
  rename('year' = 'cycle') %>% # for merging
  filter(st_cd_fips == '3901')

# merge with incumbency data
# load data
incumb_df <- read_csv("incumb_dist_1948-2022 (2).csv")
# filter for specific district and year
oh <- NA
oh <- incumb_df %>%
  filter(year == c('2018', '2020'), st_cd_fips == '3901')

# join by 'year'
oh <- left_join(oh, polls_df)

# code new incumbent variable
oh <- oh %>%
              mutate(INCUMB = case_when(winner_party == 'R' & RepStatus == 'Incumbent' ~  TRUE,
                              winner_party == 'D' & DemStatus == 'Incumbent' ~ TRUE,
                              winner_party == 'R' & RepStatus == 'Challenger' ~  FALSE,
                              winner_party == 'D' & DemStatus == 'Challenger' ~  FALSE))

# fit lm
dat_poll_inc <- oh[oh$INCUMB,]
#dat_poll_chl <- oh[!oh$INCUMB,]
mod_poll_inc <- lm(RepVotesMajorPercent ~ pct, data = dat_poll_inc)
#mod_poll_chl <- lm(DemVotesMajorPercent ~ pct, data = dat_poll_chl)

# assess model
# summary(mod_poll_inc)

## in-sample fit
mean(abs(mod_poll_inc$residuals))

plot(mod_poll_inc$fitted.values, oh$RepVotesMajorPercent,
       main="polls (incumbent)", xlab="predicted", ylab="true", 
       cex.lab=2, cex.main=2, type='n',xlim=c(40,55),ylim=c(40,55))
  text(mod_poll_inc$fitted.values, oh$RepVotesMajorPercent, oh$year)
  abline(a=0, b=1, lty=2)
  
## out-of-sample evaluation - 2018
outsamp_mod1 <- lm(RepVotesMajorPercent ~ pct, 
                   dat_poll_inc[dat_poll_inc$year != 2018,]) 
                    
outsamp_pred <- predict(outsamp_mod1, 
                        dat_poll_inc[dat_poll_inc$year != 2018,]) 

outsamp_true <- oh$RepVotesMajorPercent[oh$year == 2018]   

head(outsamp_pred - outsamp_true, 1)

## out-of-sample evaluation - 2020
outsamp_mod1 <- lm(RepVotesMajorPercent ~ pct, 
                   dat_poll_inc[dat_poll_inc$year != 2020,]) 
                    
outsamp_pred <- predict(outsamp_mod1, 
                        dat_poll_inc[dat_poll_inc$year != 2020,]) 

outsamp_true <- oh$RepVotesMajorPercent[oh$year == 2020]   

head(outsamp_pred - outsamp_true, 1)
## if we could do leave-one-out validation
# years_outsamp <- sample(oh$year, 1)
# mod <- lm(RepVotesMajorPercent ~ pct,
#           dat_poll_inc[!(dat_poll_inc$year %in% years_outsamp),])
# outsamp_pred <- predict(mod,
#                 newdata = dat_poll_inc[dat_poll_inc$year %in% 
#                                  years_outsamp,])
# mean(outsamp_pred - oh$RepVotesMajorPercent[oh$year 
#                                             %in% years_outsamp])
# 
# sqrt(mean(outsamp_pred - oh$RepVotesMajorPercent[oh$year 
#                                             %in% years_outsamp])^2)

# predict 2022
oh_22 <- polls_df %>%
  filter(st_cd_fips == '3901', year == 2022, party == 'REP')

pred_poll_inc <- predict(mod_poll_inc, oh_22, 
                          interval = "prediction", level=0.95)
pred_poll_inc

# taking it further (following 538)
# weight based on sample size

# weight based on recency

# weight based on pollster ratings

```
\pause
```{r, echo=TRUE, eval= TRUE}
# predict 2022
oh_22 <- polls_df %>%
  filter(st_cd_fips == '3901', year == 2022, party == 'REP')

pred_poll_inc <- predict(mod_poll_inc, oh_22, 
                          interval = "prediction", level=0.95)
pred_poll_inc
```


# Individual seat prediction: A district with bad polling data
## Example: Florida District 18 (something with similar demographics or likely R)
  We will "borrow" data from a district that is comparable on relevant variables to OH01. 
  Q: What do you consider relevant variables? What is a reasonable margin for comparison?
  
  General process:
      (1) Choose a district that is comparable on relevant demographic and electoral characteristics
      (2) Append polls
      (3) Consider weighting here: we probably want to attach smaller weights to the borrowed polls
      or account in some other way for the fact that they are borrowed.
      (4) Model
      (5) Predict

\begin{figure}
\includegraphics[width=0.98\textwidth]{comparison table.png}
\end{figure}

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# # find a district with low number of polls
# # table(dist_polls_2018_2022$st_cd_fips) 
# 
# # create demographic table for example
# CPVI <- c('R+5')
# VAP <- c(551000)
# black_vap <- c(0.21*551000)
# white_vap <- c(0.74*551000)
# foreignborn_vap <- c(0.04*551000)
# median_income_all <- c('$64000')
# bachelors_degree_all <- c(0.34*551000)
# urban <- c("92.5%")
# rural <- c("7.5%")
# 
# demogs_OH <- c('CPVI', 'VAP', 'black_vap','white_vap',
#             'foreignborn_vap', 'median_income_all', 
#             'bachelors_degree_all',
#             'urban', 'rural')
# var <- c(CPVI, VAP, black_vap, white_vap,
#                     foreignborn_vap,
#                     median_income_all, 
#                     bachelors_degree_all, urban, rural)
# df_oh01 <- data.frame(demogs_OH, var)
# df_oh01 
# 
# 
# # compare with FL12
# CPVI <- c('R+5')
# VAP <- c(556783)
# black_vap <- c(0.127*556783)
# white_vap <- c(0.684*556783)
# foreignborn_vap <- c(0.156*556783)
# median_income_all <- c('$68744')
# bachelors_degree_all <- c(0.34*556783)
# urban <- c("96.37%")
# rural <- c("3.63%")
# 
# demogs_FL <- c('CPVI', 'VAP', 'black_vap','white_vap',
#             'foreignborn_vap', 'median_income_all', 
#             'bachelors_degree_all',
#             'urban', 'rural')
# var <- c(CPVI, VAP, black_vap, white_vap,
#                     foreignborn_vap,
#                     median_income_all, 
#                     bachelors_degree_all, urban, rural)
# df_fl18 <- data.frame(demogs_FL, var)
# df_fl18

```
\pause

```{r, echo=TRUE, eval = TRUE, message=FALSE, warning=FALSE}
# compare
# data.frame(df_oh01, df_fl18) %>%
#   kable(format = "latex")

```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
# # load polling data
# dist_polls_2018_2022 <- read_csv("dist_polls_2018-2022.csv")
# table(dist_polls_2018_2022$st_cd_fips)
# # subset
# polls_df <- dist_polls_2018_2022 %>%
#   select(pollster, sponsors, display_name, fte_grade,
#          start_date, end_date, sample_size, population, cycle, election_date,
#          party, candidate_name, pct, st_cd_fips) %>%
#   rename('year' = 'cycle') %>% # for merging
#   filter(st_cd_fips == '1218')
# 
# # merge with incumbency data
# # load data
# incumb_df <- read_csv("incumb_dist_1948-2022 (2).csv")
# # filter for specific district and year
# fl <- NA
# fl <- incumb_df %>%
#   filter(st_cd_fips == '1218') %>%
#   filter(year == c(2020, 2018))
# 
# # join by 'year'
# fl <- left_join(fl, polls_df)
# 
# # code new incumbent variable
# fl <- fl %>%
#               mutate(INCUMB = case_when(winner_party == 'R' & RepStatus == 'Incumbent' 
#                                         & party == 'DEM'~  FALSE,
#                                         winner_party == 'R' & RepStatus == 'Incumbent' 
#                                         & party == 'REP'~  TRUE))
# # drop NAs
# fl <- fl %>%
#   drop_na(INCUMB)
# append polls from OH01

# # fit lm
# dat_poll_inc <- fl[fl$INCUMB,]
# dat_poll_chl <- fl[!fl$INCUMB,]
# mod_poll_inc <- lm(RepVotesMajorPercent ~ pct, data = dat_poll_inc)
# mod_poll_chl <- lm(DemVotesMajorPercent ~ pct, data = dat_poll_chl)
# 
# # assess model
# summary(mod_poll_inc)
# summary(mod_poll_chl)
# 
# ## in-sample fit
# mean(abs(mod_poll_inc$residuals))
# mean(abs(mod_poll_chl$residuals))
# 
# plot(mod_poll_inc$fitted.values, fl$RepVotesMajorPercent,
#        main="polls (incumbent)", xlab="predicted", ylab="true", 
#        cex.lab=2, cex.main=2, type='n',xlim=c(40,55),ylim=c(40,55))
#   text(mod_poll_inc$fitted.values, fl$RepVotesMajorPercent, fl$year)
#   abline(a=0, b=1, lty=2)
#   
# ## out-of-sample evaluation - 2018
# outsamp_mod1 <- lm(RepVotesMajorPercent ~ pct, 
#                    dat_poll_inc[dat_poll_inc$year != 2018,]) 
#                     
# outsamp_pred <- predict(outsamp_mod1, 
#                         dat_poll_inc[dat_poll_inc$year != 2018,]) 
# 
# outsamp_true <- fl$RepVotesMajorPercent[oh$year == 2018]   
# 
# head(outsamp_pred - outsamp_true, 1)
# 
# ## out-of-sample evaluation - 2020
# outsamp_mod1 <- lm(RepVotesMajorPercent ~ pct, 
#                    dat_poll_inc[dat_poll_inc$year != 2020,]) 
#                     
# outsamp_pred <- predict(outsamp_mod1, 
#                         dat_poll_inc[dat_poll_inc$year != 2020,]) 
# 
# outsamp_true <- fl$RepVotesMajorPercent[oh$year == 2020]   
# 
# head(outsamp_pred - outsamp_true, 1)
# ## if we could do leave-one-out validation
# # years_outsamp <- sample(oh$year, 1)
# # mod <- lm(RepVotesMajorPercent ~ pct,
# #           dat_poll_inc[!(dat_poll_inc$year %in% years_outsamp),])
# # outsamp_pred <- predict(mod,
# #                 newdata = dat_poll_inc[dat_poll_inc$year %in% 
# #                                  years_outsamp,])
# # mean(outsamp_pred - oh$RepVotesMajorPercent[oh$year 
# #                                             %in% years_outsamp])
# # 
# # sqrt(mean(outsamp_pred - oh$RepVotesMajorPercent[oh$year 
# #                                             %in% years_outsamp])^2)
# 
# # predict 2022
# fl_22 <- polls_df %>%
#   filter(st_cd_fips == '1218', year == 2022, party == 'REP')
# 
# pred_poll_inc <- predict(mod_poll_inc, fl_22, 
#                           interval = "prediction", level=0.95)
# pred_poll_inc
# 
# # taking it further (following 538)
# # weight based on sample size
# 
# # weight based on recency
# 
# # weight based on pollster ratings


```

# Aggregating seat predictions

As we just saw, predicting individual seats involves a lot of discrete work. 

See Yao's code in the Class dropbox for starter code on how to do aggregate our seat-level predictions.

# The incumbency advantage

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(readr)
incumb_df <- read_csv("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/fundamentals II_ incumbency/incumb_dist_1948-2020.csv") # replace with your wd()

incumb_df <- 
  incumb_df %>%
  mutate(winner_candidate = case_when(winner_party == 'R' ~ 'RepCandidate',
                                      TRUE ~ 'DemCandidate'),
         winner_candidate_inc = case_when(winner_candidate == 'RepCandidate' & RepStatus == 'Incumbent' ~ "Incumbent",
                                                 winner_candidate == 'RepCandidate' & RepStatus == 'Challenger' ~ "Challenger",
                                             winner_candidate == 'DemCandidate' & DemStatus == 'Incumbent' ~ "Incumbent",
                                             winner_candidate == 'DemCandidate' & DemStatus == 'Challenger' ~ "Challenger",
                                             TRUE ~ "Challenger"))
```

## The incumbency advantage: descriptive statistics

How many post-war elections where \textcolor{purple}{incumbent candidate won?} \pause
\footnotesize
```{r, warning=FALSE, message=FALSE}
# in how many elections did the incumbent candidate win
incumb_df %>%
  select(year, winner_party, winner_candidate, winner_candidate_inc) %>%
  mutate(winparty_last = lag(winner_party, order_by = year),
        wincand_last  = lag(winner_candidate,  order_by = year),
        wincand_last_inc  = lag(winner_candidate_inc,  order_by = year)) %>%
  mutate(reelect.cand = wincand_last_inc == winner_candidate_inc) %>%
  filter(year > 1948) %>%
  group_by(reelect.cand) %>% 
  summarise(n = n()) %>% 
  as.data.frame() %>%
  kable(format = "latex")

```

## The incumbency advantage: descriptive statistics

How many post-war elections where \textcolor{purple}{incumbent party won?}
\footnotesize
```{r, warning=FALSE, message=FALSE}
incumb_df %>%
  select(year, winner_party, winner_candidate, winner_candidate_inc) %>%
  mutate(winparty_last = lag(winner_party, order_by = year),
        wincand_last  = lag(winner_candidate,  order_by = year)) %>%
  mutate(reelect.party = winparty_last == winner_party) %>%
  filter(year > 1948) %>%
  group_by(reelect.party) %>% 
  summarise(n = n()) %>% 
  as.data.frame() %>%
  kable(format = "latex")

```

<!--
Comports with what see saw in the Lecture video: incumbents often win, yet the incumbent party is still punished
-->

## So, what's the deal with incumbency?

\textbf{Some incumbency advantages}: \pause

  * More media coverage
  * Campaign finance access \pause
  * \textbf{\textcolor{pink}In the presidential context, {"Pork"}} $\rightsquigarrow$ short-term economic gains (Bartels 2008), 
  credit-claiming (Kriner and Reeves 2012)

\textbf{Some incumbency (dis/non-)advantages}: \pause

  * Polarized electorate $\rightsquigarrow$ partisanship, not incumbency matters (Donovan et al. 2019)
  * Recessions, disasters $\rightsquigarrow$ blame attribution (Achen and Bartels 2016)
  * Incumbency fatigue

## (2) Credit and blame: the time-for-change model
\pause 
Alan Abramowitz's \textbf{time-for-change} model is a classic model of incumbency
and, since 1992, has a \underline{true} out-of-sample PV prediction error of $1.7\%$ 
\textcolor{red}{(Typically used in the presidential forecasting context, but if we consider the fact 
that midterm elections are often referenda on the president, it may be relevant).}

$$
\underbrace{\texttt{pv2p}}_{incumbent \ party} = A + B_1 \underbrace{\texttt{G2GDP}}_{Q2 \ \ GDP \ growth} + B_2 \underbrace{\texttt{NETAPP}}_{Gallup \ job \ approval} + B_3 \underbrace{\texttt{TERM1INC}}_{sitting \ pres}
$$
\tiny
\begin{center}
\href{https://pollyvote.com/en/components/models/retrospective/fundamentals-plus-models/time-for-change-model/}{(\texttt{pollyvote.com} model repo)}
\end{center}
\normalsize

# Preview of next week: Probabilistic models

## One major problem with linear regression

* When we fit a linear regression model $Y = \alpha + \beta X$, there are no restrictions on $Y$. What's wrong with that? \pause

* $\rightsquigarrow$ It is possible to have a prediction interval lower bound $< 0$ (\textbf{out of support}). \pause

* \textbf{This often occurs when we are extrapolating but also when there is sparse data} \pause \textbf{(e.g. district-level polls).}

## Solution: probabilistic models

* In a linear regression, $$DemPV_{state} = \alpha + \beta_1 x_1 + \ldots + \beta_k x_k,$$ our probabilistic assumption is that errors in predicted PV follow a bell curve, $DemPV_{state} - \widehat{DemPV_{state}} \sim Normal()$ \pause

* In reality, the process of elections is some "draw" of voters from the voter-eligible population (VEP) turning out to vote for a party: \small
$$
\Pr\Big(DemPV_{district} = \texttt{2  million} \ | \ VEP_{district} = \texttt{5  million}\Big) = 
f(\alpha + \beta_1 x_1 + \ldots + \beta_k x_k)
$$ 
with some other probabilistic assumption, $DemPV_{district} \sim ?$. \pause

* A model that allows the DV or error to have a non-normal distribution (specified by a particular choice of function $f(\cdot)$) is called a \textbf{generalized linear model} $\rightsquigarrow$ \textcolor{red}{more on this and how to apply them in \textsf{R} next time!}

## Blog Extensions
\tiny
1. \textbf{How accurate are expert predictions? pt.1} Visualize actual voteshare (seatshare) in 2018 and compare that to various expert predictions for that election cycle. How do they compare? 

2. \textbf{How accurate are expert predictions? pt.2} Visualize actual voteshare (seatshare) in 2018 and compare that to various expert predictions for that election cycle. How do they compare? Create 3 maps: (1) a map that visualizes voteshare (seatshare) at the district-level; (2) a map that visualizes expert predictions at the district-level; (3) a map that visualizes the difference between actual voteshare and expert prediction at the district-level.

This is going to require you to use your own discretion in coding up variables.

On (2): these expert predictions are in the form of "lean D/R," "likely D/R," etc. Transform these variables into a continuous numeric variable, i.e. "Likely R" ~ -2, "Lean R" ~ -1, "Tossup" ~ 0, "Lean D" ~ 1, "Likely D" ~ 2, and so on. Visualize the results.

On (3): you will need to figure out how to compare voteshare and expert predictions. One possibility: transform one of the variables to be on the same scale as the other variable. Ex: voteshare of 54% for Democrats ~ "Safe D," voteshare of 52% for Democrats ~ "Likely D," etc. or vice versa, "Safe D" district ~ "54% voteshare" ...

