---
title: "Election Administration"
subtitle: "Gov 1347: Election Analytics"
author: Kiara Hernandez
date: \today
institute: Harvard University
fontsize: 10pt
output:
 beamer_presentation:
    keep_tex: true
    theme: metropolis
    latex_engine: pdflatex
    slide_level: 2
    highlight: zenburn
    incremental: false
header-includes:
  - \setbeamercolor{frametitle}{bg=orange}
  - \hypersetup{colorlinks,citecolor=orange,filecolor=red,linkcolor=brown,urlcolor=blue}
---

## Today's agenda

* **Why do we care about election admin?**

* **Review: What do we know about the cost of voting (from discussion)?**
    * Fundamental questions
      - What is the relationship between the cost of voting index and turnout? Vote margin by party?
      - How do we \textit{empirically} understand the causal relationship between COVI and election outcomes?
    * How this might affect election night
    
* **Redistricting: What to do about it in our models**
    * Fundamental questions

      -How does (can) it substantively affect election outcomes?
    * Suggestions for incorporating it into a predictive model
    
* **Updating our data**
    * Pulling up-to-date polling and expert ratings
    * Pulling 2022 district shapefiles from the Census

# Why do we care about election admin?

## What is election administration?

\begin{table}
\begin{tabular}{r|l}
Component & Examples \\
\hline
\textbf{Information} & registration status, \\[-0.2em]
& polling place look-up \\
\textbf{Registration} & eligibility, roll-off \\
\textbf{Convenience voting} & {vote-by-mail}, early voting \\
\textbf{Ballot access} & wait times, photo ID, \\[-0.2em]
& provisional balloting \\
\textbf{Ballot count} & security, technology,  \\[-0.2em]
& audits \\
\textbf{News coverage} & exit polls, decision desks \\
\textbf{Redistricting} & polling places, \textcolor{red}{incentives to vote (individual power)}
\end{tabular}
\end{table}

## Discussion: what do we know about the cost of voting?

-From discussion: 

$U = pB - C + D$, where 
-$U$ = Utilty \newline
  -if $U$ > 0, vote \newline
  -if $U$ < 0, don't vote \newline
-$p$ = probability of casting pivotal vote \newline
-$B$ = benefit of candidate winning \newline
-$C$ = cost \newline
-$D$ = expressive benefit of voting

## COVI and turnout - visualization
\tiny
-What time trends do we see? \newline
-What regional and state trends do we see?

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(sf)  ##'simple features'
library(tmap) #for maps!
library(tmaptools) #tmap extensions
library(spData) ##data for example

# master data
Voting1996 <- read_excel("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/laws and election admin/data/Master 1996 2022 Election voting requirements.xlsx", sheet = "1996")
Voting2000 <- read_excel("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/laws and election admin/data/Master 1996 2022 Election voting requirements.xlsx", sheet = "2000")
Voting2004 <- read_excel("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/laws and election admin/data/Master 1996 2022 Election voting requirements.xlsx", sheet = "2004")
Voting2008 <- read_excel("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/laws and election admin/data/Master 1996 2022 Election voting requirements.xlsx", sheet = "2008")
Voting2012 <- read_excel("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/laws and election admin/data/Master 1996 2022 Election voting requirements.xlsx", sheet = "2012")
Voting2016 <- read_excel("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/laws and election admin/data/Master 1996 2022 Election voting requirements.xlsx", sheet = "2016")
Voting2020 <- read_excel("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/laws and election admin/data/Master 1996 2022 Election voting requirements.xlsx", sheet = "2020")
Voting2020covid <- read_excel("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/laws and election admin/data/Master 1996 2022 Election voting requirements.xlsx", sheet = "2020CCOVI")
Voting2022 <- read_excel("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/laws and election admin/data/Master 1996 2022 Election voting requirements.xlsx", sheet = "2022")

VotingDataMaster <- bind_rows(Voting1996,Voting2000,Voting2004,Voting2008,Voting2012,Voting2016,Voting2020,Voting2022)

VotingDataMaster <- VotingDataMaster %>%
  mutate(OnlineRegistration = ifelse(NOonlineregistration == 0,'Yes','No'),
         OnlineRegistration = ifelse(is.na(OnlineRegistration)==T,'No',OnlineRegistration),
         HourWait = ifelse(ExcessiveWaitTime ==1, 'Yes','No'))

# rankings
covi_1996_2022 <- read_excel("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/laws and election admin/data/COVI Values 1996-2022 website addl.xlsx") %>%
  mutate(FinalCOVI = ifelse(year == 2022, COVI2022_IRT,FinalCOVI),
         FinalRank = ifelse(year == 2022, COVI2022_IRTrank,FinalRank))%>%
  select(state,year,FinalCOVI,FinalRank)

# attach geographic data
us_states <- us_states%>%
  mutate(state = state.abb[match(NAME,state.name)])%>%
  filter(GEOID!=11) #remove DC

covi_1996_2022 <- left_join(us_states,
                       covi_1996_2022,
                       by = 'state')

VotingDataMaster <- left_join(us_states,
                       VotingDataMaster,
                       by = c('state' = 'State'))
# merge in turnout dfs
# cvap
library(readr)
cvap_df <- read_csv("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/laws and election admin/data/cvap_district_2012-2020_clean.csv")
# cvap_df <- cvap_district_2012_2020_clean
# district-level votes
dist_pv_df <- read_csv("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/fundamentals II_ incumbency/incumb_dist_1948-2020 (3).csv")

# select relevant years from voting data
dist_pv_df <- dist_pv_df %>%
  filter(year == 2012 | year == 2014 | year == 2016 | year == 2018 | year == 2020)

# mutate geoid for merging
cvap_df <- cvap_df %>%
  dplyr::rename(st_cd_fips = geoid)

# merge
dist_pv_cvap <- dist_pv_df %>%
  inner_join(cvap_df, by = c('st_cd_fips', 'year'))

# mutate turnout
dist_pv_cvap <- dist_pv_cvap %>%
  mutate(totalvotes = RepVotes + DemVotes,
         turnout = totalvotes/cvap)

# merge in demographic data
demog_df <- read_csv("~/Dropbox/ElectionAnalytics_Midterms/Lab sessions/laws and election admin/data/demographic_2009_2020.csv")

# demographic surges
demog_df <- demog_df %>% 
  # change age var name
   dplyr::rename("age20_29" = "20_29", 
                "age30_44" = "30_44",
               "age45_64" = "45_64", 
                "age65" = "65+",
               "hispanic" = "hispanic or latino",
               "indigenous" = "native american")

demog_df <- demog_df %>%
    group_by(st_cd_fips) %>%
    mutate(Asian_change = asian - lag(asian, order_by = year),
           Black_change = black - lag(black, order_by = year),
           Hispanic_change = hispanic - lag(hispanic, order_by = year),
           Indigenous_change = indigenous - lag(indigenous, order_by = year),
           White_change = white - lag(white, order_by = year),
           Female_change = female - lag(female, order_by = year),
           Male_change = male - lag(male, order_by = year),
           age20_change = age20_29 - lag(age20_29, order_by = year),
           age3045_change = age30_44 - lag(age30_44, order_by = year),
           age4565_change = age45_64 - lag(age45_64, order_by = year),
           age65_change = age65+ - lag(age65, order_by = year)
           )

# merge surges with turnout, pv
dist_pv_cvap_demog <- dist_pv_cvap %>%
  inner_join(demog_df, by = c('st_cd_fips', 'year'))

# merge with original COVI rankings - on state, not district
# rename state
covi_1996_2022 <- covi_1996_2022 %>%
  dplyr::rename(state_abb = state, state = NAME)

demog_covi_df <- merge(covi_1996_2022, dist_pv_cvap_demog, by = c('state', 'year'))

# clean it up - select relevant variables
demog_covi_df <- demog_covi_df %>%
  select('state', 'year', 'REGION', 'AREA', 'state_abb', 'FinalCOVI', 'FinalRank', "winner_party",
         "RepVotes","RepCandidate","RepStatus","DemVotes","DemCandidate","DemStatus",
         "RepVotesMajorPercent", "DemVotesMajorPercent", "winner_candidate","winner_candidate_inc", "st_cd_fips", "president_party",
         "cvap", "totalvotes","turnout", "male","female","age20_29","age30_44","age45_64","age65","white","black",
         "indigenous","asian","pacific islander","other","mixed","hispanic",
         "Asian_change","Black_change","Hispanic_change","Indigenous_change","White_change","Female_change",
         "Male_change","age20_change","age3045_change","age4565_change","age65_change")
# create state-level variables to match COVI data
# state-level turnout // state level votes and cvap
demogS_covi_df <- demog_covi_df %>%
  group_by(state, year) %>%
  # dplyr::mutate(avg_state_turnout = mean(turnout)) %>%
  dplyr::summarise(avg_state_turnout = mean(turnout),
            st_dem_votes = sum(DemVotes),
            st_cvap = sum(cvap)) %>%
  as.data.frame() %>%
  select(-geometry)
# remerge with original df
demog_covi_df <- merge(demog_covi_df, demogS_covi_df, by = c('state', 'year'))

## visualization ##
# covi x turnout - 2012
# install.packages('ggh4x')
library(ggh4x)
plot <- ggplot(demog_covi_df,
        mapping = aes(x = FinalRank, y = avg_state_turnout,
       label = state_abb, colour = REGION))  + 
  #facet_grid(cols = vars(year), scales = "free") + 
  facet_wrap(~ year, scales='free_x') +
  geom_point() +
  #theme(panel.spacing = unit(2, "lines")) +
  #force_panelsizes(cols = c(0.3, 1)) +
  geom_text(hjust=0, vjust=0) +
  geom_smooth(method = "lm", color = "red") +
  theme(legend.position = "bottom")
```
\tiny
```{r, echo=TRUE, eval=TRUE, warning=FALSE}
plot
```


```{r, eval=TRUE, echo=FALSE, warning=FALSE}
# linear regression
fit1 <- lm(avg_state_turnout ~ FinalRank, data = demog_covi_df)
# with controls
fit2 <- lm(avg_state_turnout ~ FinalRank + black + age20_29, data = demog_covi_df)
```

## COVI and turnout - regression

```{r, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, results='asis'}
library(stargazer)
stargazer(fit1, fit2, header=FALSE, type='latex', no.space = TRUE,
          column.sep.width = "3pt", font.size = "scriptsize", single.row = TRUE,
          keep = c(1:7, 62:66), omit.table.layout = "sn",
          title = "State-level turnout and COVI")
```


## COVI and voteshare margin by party - visualization

```{r, echo=FALSE, eval=TRUE, warning=FALSE}
# state-level average dem voteshare
demog_covi_df <- demog_covi_df %>%
  group_by(state, year) %>%
  dplyr::mutate(avg_state_demshare = mean(DemVotesMajorPercent))

# visualization
plot2 <- ggplot(demog_covi_df,
        mapping = aes(x = FinalRank, y = avg_state_demshare,
       label = state_abb, colour = REGION))  + 
  facet_wrap(~ year, scales='free_x') +
  geom_point() +
  #theme(panel.spacing = unit(2, "lines")) +
  #force_panelsizes(cols = c(0.3, 1)) +
  geom_text(hjust=0, vjust=0) +
  geom_smooth(method = "lm", color = "red") +
  theme(legend.position = "bottom")
```
```{r, echo=TRUE, eval=TRUE, warning=FALSE}
plot2
```

```{r, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
# linear regression
fit3 <- lm(avg_state_demshare ~ FinalRank, data = demog_covi_df)
# with controls
fit4 <- lm(avg_state_demshare ~ FinalRank + black + age20_29, data = demog_covi_df %>%
             filter(avg_state_demshare != 0 | avg_state_demshare != 1))
```

## COVI and voteshare margin by party - regression

```{r, eval=TRUE, echo=TRUE, warning=FALSE, results='asis'}
stargazer(fit3, fit4, header=FALSE, type='latex', no.space = TRUE,
          column.sep.width = "3pt", font.size = "scriptsize", single.row = TRUE,
          keep = c(1:7, 62:66), omit.table.layout = "sn",
          title = "State-level Democratic voteshare and COVI")
```

## Prediction using COVI - can we predict the probability of Democrats winning in 2022 from historical COVI data?
```{r, echo=FALSE, eval=TRUE, warning=FALSE}
# training data - 1996-2018 (leaving out 2022)
# define years
yrs <- c(2012, 2014, 2016, 2018, 2020)
train_covi_df <- demog_covi_df %>%
  filter(year %in% yrs) %>%
  group_by(state) %>% 
  group_nest() %>% 
  mutate(data = map(data, ~unnest(., cols = c())))
  
# test data - 2022
test_covi_df <- covi_1996_2022 %>%
  filter(year == 2022) %>%
  group_by(state) %>% 
  group_nest() %>%
  mutate(data = map(data, ~unnest(., cols = c())))

# fit training model - state-level 
models <- train_covi_df %>% 
  mutate(model = map(data, ~glm(cbind(st_dem_votes, st_cvap-st_dem_votes) ~ FinalRank, 
                                data = .x, family = binomial))) %>% 
  select(-data)

# extract model results
model_results <- models %>% 
  mutate(mcf_r_squared = map_dbl(model, ~with(summary(.x), 1 - deviance/null.deviance)))

# predict 2022 (with intervals!)
pred_2022 <- test_covi_df %>%
  # inner join as there may not be historical models for some districts
  inner_join(models, by = c("state")) %>% 
  mutate(preds = map(.x = model, .y = data, ~predict(object = .x, newdata = as.data.frame(.y),
                                                     se.fit = TRUE,
                                                     type = "response")),
         lower = map_dbl(.x = preds, ~.x$fit - (2 * .x$se.fit)),
         fitted = map_dbl(.x = preds, ~.x$fit),
         upper = map_dbl(.x = preds, ~.x$fit + (2 * .x$se.fit))) %>%
  select(state, lower, fitted, upper)

st <- covi_1996_2022 %>%
filter(year == 2022) %>%
select(state, FinalRank) %>%
filter(state != 'Rhode Island') %>%
  as.data.frame()
# merge
pred_2022 <- merge(pred_2022, st, by = 'state')

ggplot(pred_2022, aes(FinalRank, y=fitted)) + 
  geom_point(alpha=.5) +
  stat_smooth(method="glm", se=FALSE, method.args = list(family=binomial))
```

## Prediction using COVI - can we predict turnout from COVI?

-Try this on your own! Would this be a more or less useful than predicting turnout based on historical turnout data?

## Trying to empirically understand the relationship between election admin and voting

-Read the following sections of this paper by Grimmer et al. (2018):
  -INTRO; USE OF NATIONAL SURVEYS FOR STATE RESEARCH; ESTIMATING VOTER ID LAWS’ EFFECTS ON TURNOUT

-In small groups, discuss the following questions:

1) What makes identifying the effects of voter ID laws (and election admin more broadly) on turnout empirically challenging? (Think endogeneity)

2) What analyses do the authors present as evidence?

3) What should we take away normatively from a paper like this? 

## The (maybe not so bad after all) bane of our existence as Midterm election forecasters: redistricting
\tiny
-High-level thoughts:

(1) In general, we can assume that most districts retain a substantial portion of their previous voters from one redistricting cycle to the next

  -We want to corroborate this, though — but how? $\rightarrow$ let's take a look at district-level population overlap between cycles
  
  -Courtesy of Chris Kenny from the Gov PhD program: '../section data/pop_overlap_vtd_estimates.csv' \newline
    -What this tells us: an estimated x% of new district i in 2022 came from old districts {i,j,k} in 2012... \newline
    -Population and VAP come from the 2020 Decennial Census \newline
    -Take the map outputs from the 50-states simulations and gets a VTD (voting district) level estimate \newline
      -https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/SLCD3E \newline

  -What would be the best way to use this? No right answer, but a few suggestions:\newline
  (a) interact the proportion overlap with the old district with IVs \newline
    ex: interact with incumbency. This would dampen the effect of incumbency. In any interaction, you would basically being downweighting or upweighting accordingly. \newline
    -Assess model fit with and without interactions \newline
    
## Redistricting options continued

(2) Reconstructing the dependent variable \newline
  -Recalculate new vote totals from historical vote totals at the precinct level for new districts $\rightarrow$ 
  Lucy's VTD estimates! \newline
  -Precinct boundaries do change some, so this approach does involve some error \newline
  
(3) Including a variable at the district level that accounts for changes in expert predictions, so if lean R/D doesn’t change between two districts from one redistricting cycle to the next, we may feel more comfortable making predictions based on last redistricting cycles' data \newline

(4) Because we're using all districts 2012-2020 as the training data (as opposed to only one district 2012-2020) and applying national betas to the individual seat, we can feel more comfortable with our predictions \newline

